Moral Foundations of U.S. Political News Organizations

### 2019-04-14 pilot_manuscript.Rmd
Just started Conclusion. More to come...

Table \@ref(tab:exp2-source-descriptives-kav) displays the descriptive statistics for sources' writing on the Kavanaugh confirmation. The sources are similar in some aspects yet different in others. Valence appears to be slightly positive across all sources. The standard deviations seem to indicate little to no presence of a difference in valence. *The New York Times* published the greatest number of articles as well as total words. *Breitbart* featured the lowest number of articles, and *NPR* the lowest number of total words from all articles. Per indvidual article, however, *Breitbart* appears to feature the highest average number of words as well as unique words. Once again the standard deviations call into question any apparent differences between sources. Finally, *Fox News* appears to be the most readable source on average, while *The New York Times* is the least readable. This might be attributable to the greater number of tokens in the average *New York Times* article compared to *Fox News*. As before, the standard deviations for readability hamper one's ability to draw solid conclusions regarding differences in the sources' use of language. Table \@ref(tab:exp2-source-descriptives-gs) displays descriptive statistics for articles about the partial government shutdown.

@@

```{r exp2-source-descriptives-gs, results = 'asis', echo = FALSE, message = F}
#Kav
#GS
final_gs$uniquewords = NA
final_gs$valenceavg = NA

@@

@@

@@

>>>>

Table \@ref(tab:exp2-source-descriptives-kav) displays the descriptive statistics for sources' writing on the Kavanaugh confirmation hearing. The sources were similar in most basic linguistic aspects, except for number of articles. Sources varied on the number of articles published pertaining to this topic. For example, *The Huffington Post* appears to have published only 27 aritcles while *NPR* published 757 articles. Valence was found to be slightly positive across all sources with the standard deviations precluding any conclusions supporting the presence of differences. *Slate* produced the most total words with the most tokens on average. *Rush Limbaugh* featured the highest number of types on average. *The Blaze* was the most readable source on average by grade level while *Fox News* was the least readable. The standard deviations for these statistics, however, preclude conclusions regarding differences in the sources' use of language.

Table \@ref(tab:exp2-source-descriptives-gs) displays descriptive statistics for articles about the partial government shutdown. Like the Kavanaugh hearing, the sources were similar in average valence (slightly positive). Once again, there was variation in the number of articles published by each sorce on this topic. *The Hugginton Post* and *The Blaze* published fewer than 100 articles while *Slate* published 1,013 articles. *Slate* again featured the most total words and mean tokens. *Rush Limbaugh* again had the most types on average. For this topic, *Slate* was the most readable while *The Huffington Post* was the least readable. For each statistic, the standard deviations render any assertions regarding linguistic differences inconclusive, at least on a descriptive level.

@@

```{r exp2-source-descriptives-gs, results = 'asis', echo = FALSE, message = F}
#GS
final_gs$uniquewords = NA
final_gs$valenceavg = NA

@@

To analyze if news sources adhered to differences in word use based on their target audience, we utilized a multilevel model (MLM) to analyze if news sources leveraged different vocabularies based on target audience. Once again, the researchers used the *nlme* library in *R* [@Pinheiro2017]. Each foundation's weighted percentage was predicted by the source's political lean, using the individual source as a random intercept to control for the nested ata structure. MLMs were constructed from datasets compiled for both topics of interest: the Kavanaugh hearing and the partial government shutdown of 2018-2019.

For the Kavanuagh topic, the multilevel model indicated the presence of a significant effect for *harm/care*, but the practical effect denoted by Cohen's *d* was found to be small. There were no other significant or practical effect of political lean for any of the other four moral foundations. The effect for *harm/care* was in the hypothesized direction with liberal sources tending to use more positively *harm/care* words than consevative sources. Descriptive and test statistics, *p*-values and effect sizes (Cohen's *d*) can be found in Table \@ref(tab:exp2-tablekav).

For news articles about the partial U.S. Federal Government Shutdown of 2018-2019, there were no significant or practical effects of political lean for the moral foundations. A small-to-medium effect size was observed for *authority/respect*. The effect was in the predicted direction as well, as conservative sources tended to offer more positive endorsements of the foundation than liberal sources. This is similar to Experiment 1 in which the largest effect size was observed for *authority/respect*. As noted before, the effect found in Experiment 1 was in the opposite direction as what was hypothesized. Thus, it is difficult to draw comparisons between the two studies despite the similar pattern for effect size. Owing to a lack of matching effects for either Experiment 1 or the Kavanaugh topic, there is doubt as to whether or not a practical or generalized effect exists for *authority/respect*.

Based on the weighted percent values for the five foundations applied to both topics, MFD words seem to make up little of the article text. A similar pattern was observed for the results in Experiment 1. As in Experiment 1, the percentages and means seem to indicate a generally positive endorsement of all five moral foundations across both political leanings.

@@

The results obtained in Experiment 2 did not completely confirm either of the hypotheses. The researchers found little compelling evidence of an effect of partisan lean on MFD endorsement. The only significant effect was found for *harm/care* in the Kavanuagh confirmation hearing dataset. While the effect was in the hypothesized direction (higher positive endorsement for liberal sources), the effect size was small. While there were no significant effects found for the government shutdown topic, a small-to-medium effect size was calculated for the *authority/respect* foundation. Once again, the effect was in the hypothesized direction with conservative sources providing higher postive endorsements of the foundation relative to liberal sources.

Both of these effects are consistent with both the resesarch hypotheses as well as previous findings in Moral Foundations Theory scholarship. It should be noted these effects were found in isolation, and similar patterns were not found across the other studies. These findings undermine any inferences of political bias stemming from the content of news articles. The results obtained for Experiment 2 call into question the Moral Foundations Dictionary's efficacy as a tool for establishing differences between political news sources by partisan lean.

@@

Within the theoretical framework of Moral Foundations Theory [@Haidt2007], the researchers attempted to devise a method leveraging the Moral Foundations Dictionary [@Graham2009] in order to quantify political bias stemming from content published by several prominent American news sources. This method was carried out in several parts. First, the researchers collected the articles through web scraping techniques in *R* [@Wickham2016]. Second, the text was processed using *ngram* and *tm* in *R* to render it in a form conducive to computerized statistical analysis [@Feinerer2017]. Third, percent occurrence of words in the Moral Foundations Dictionary (MFD) was calculated on each source in order to establish how often words indicating endorsement of the five moral foundations appeared. Frequency of MFD word occurrence per article was weighted for valence by multiplying the scores by mean *z-scored* valence ratings from @Warriner2013. Finally, the researchers constructed multi-level models using *nlme* in *R* [@Pinheiro2017] for each foundation in which weighted percentages were predicted by each source's political lean. Individual source was held as a random intercept to control for correlated error. This process was undertaken once in Experiment 1 for general political news and twice in Experiment 2 using articles pertianing to two specific topics in American news: Brett Kavanaugh's confirmation to the Supreme Court and the most recent (as of this writing) partial U.S. Government shutdown from December 2018 to January 2019.

In Experiment 1, the researchers analyzed general political news articles from four prominent U.S. news sources: *The New York Times*, *NPR*, *Fox News*, and *Breitbart*. They hypothesized *The New York Times* and *NPR* would endorse moral foundations representative of liberal political views and that *Fox News* and *Breitbart* would endorse foundations indicative of conservative views. The researchers assigned a political lean (liberal or conservative) to the sources based upon consumer preferences established by @Mitchell2014. After collecting the articles through web scraping, processing the article text, calculating percent occurrence of MFD words, and weighting that value by valence, the researchers constructed the MLM. The results obtained were not significant in any statistical or practical sense. There was a small effect found for the *authority/respect* foundation according to Cohen's *d*, but the effect suggested endorsement for that foundation opposite to what was hypothesized. The researchers viewed these results as an opportunity to alter their method for use in Experiment 2.

In Experiment 2, the researchers adjusted their method in order to increase their potential ability to quantify political bias in U.S. news sources. First, the researchers expanded the number of news sources from which articles were collected. They added *Slate*, *Politico*, and *Huffington Post* to the list of liberal sources and *The Blaze*, *Rush Limbaugh*, and *Sean Hannity* to the conservative sources. Once again, political bias groups were assigned based on @Mitchell2014's findings. Second, they elected to collect articles and compile datasets pertaining to two specific topics in U.S. news: the Kavanaugh hearing and the government shutdown. These topics were chosen due to their relevance to multiple moral foundations. Other than those two alterations, the same method (scraping, processing, weighting, and multi-level modelling) was employed for Experiment 2.

For articles pertaining to the Kavanaugh hearing, there was a statistically significant effect for *harm/care*, but the effect size was modest. No other statistically or practically significant effects were found for the Kavanaugh topic. Likewise, there were no statistically or pracitically signifcant effects revealed for the government shutdown topic. However, there was a small-to-medium effect size for *authority/respect*. The fact these patterns do not appear in any of the other studies undertaken herein decreases the researchers' ability to draw any valuable conclusions regarding the MFD's ability to reveal political bias in American news media.
